---
title: "Cure-Fraction Survival Trees and Forests with cureforest"
author: "Pedro Luiz Ramos and Lucas Osses"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cure-Fraction Survival Trees and Forests with cureforest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope

`cureforest` is designed for right-censored studies in which the scientific
target is a long-term event-free or cure fraction. The target is a
population-level probability inferred from the survival tail. It is not an
observed subject-level cure label, and it requires adequate follow-up.

The package supplies:

1. `survcure()`, an honest cure-fraction survival tree;
2. `randomforestcure()`, a randomized ensemble of honest trees;
3. matched log-rank splitting through `splitrule = "logrank"`;
4. population, susceptible, and dynamic-cure predictions.

## Simulated long-term survival data

The following small example has one variable that changes the cure fraction
and one variable that changes event timing among susceptible subjects.

```{r data}
library(survival)
library(cureforest)

set.seed(2401)
n <- 600
x_cure <- runif(n)
x_latency <- factor(rbinom(n, 1, 0.5), labels = c("slow", "fast"))
noise <- rnorm(n)

pi0 <- plogis(-0.8 + 3.0 * (x_cure - 0.5))
long_term <- rbinom(n, 1, pi0)
rate <- ifelse(x_latency == "fast", 0.35, 0.14)
susceptible_time <- pmin(rexp(n, rate), 9)
event_time <- ifelse(long_term == 1, Inf, susceptible_time)
censor_time <- runif(n, 0, 12)

dat <- data.frame(
  time = pmin(event_time, censor_time),
  status = as.integer(event_time <= censor_time),
  x_cure = x_cure,
  x_latency = x_latency,
  noise = noise
)
```

## Fit an honest tree

The 7--9 year window is specified before fitting. Both daughter nodes must
retain adequate risk-set support in this interval.

```{r tree}
tree <- curetree(
  Surv(time, status) ~ x_cure + x_latency + noise,
  data = dat,
  tail.start = 7,
  tail.end = 9,
  nodesize = 30,
  min.tail.at.risk = 8,
  min.tail.end.at.risk = 3,
  maxdepth = 2,
  nsplit = 8,
  seed = 11
)
tree
summary(tree)
```

The estimated cure probability and population survival are available through
the standard `predict()` generic.

```{r tree-predict}
predict(tree, dat[1:6, ], type = "cure")
predict(tree, dat[1:3, ], type = "survival", times = c(3, 6, 9))
```

## Fit a cure-fraction forest

This vignette uses 20 trees for speed. Analyses ordinarily use a larger
ensemble and should examine stability across random seeds.

```{r forest}
forest <- cureforest(
  Surv(time, status) ~ x_cure + x_latency + noise,
  data = dat,
  ntree = 20,
  mtry = 2,
  nodesize = 25,
  maxdepth = 2,
  nsplit = 8,
  tail.start = 7,
  tail.end = 9,
  min.tail.at.risk = 7,
  min.tail.end.at.risk = 3,
  n.cores = 1,
  seed = 22
)
forest
cure_importance(forest)
```

```{r forest-predict}
pred <- predict(forest, dat[1:4, ], type = "all", times = c(3, 6, 9))
pred$cure
pred$survival
pred$susceptible
pred$dynamic
```

## Matched log-rank comparison

Use the same tree controls with `splitrule = "logrank"` to isolate the effect
of the splitting target. A generic predictive advantage should not be assumed:
the comparison asks whether cure-oriented splitting creates different,
tail-supported strata.

```{r matched-comparison}
logrank_tree <- curetree(
  Surv(time, status) ~ x_cure + x_latency + noise,
  data = dat,
  splitrule = "logrank",
  tail.start = 7,
  tail.end = 9,
  nodesize = 30,
  min.tail.at.risk = 8,
  min.tail.end.at.risk = 3,
  maxdepth = 2,
  nsplit = 8,
  seed = 11
)

summary(tree)$splits
summary(logrank_tree)$splits
```

## Reporting recommendations

Report the prespecified tail window, numbers at risk at its beginning and end,
terminal fallback frequency, and sensitivity to alternative scientifically
reasonable windows. Prediction metrics complement these diagnostics but do not
replace them.
